---
title: "Description of Subjects"
author: "Ryan Zhou"
date: "2023-11-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description of Subjects
```{r}
data <- read.csv("/Users/ryanzhou/Documents/Study/Cornell/2023/Course/STSCI 5160/Prelim/STSCI-4110-5160-Prelim2/Handout/final_cardiac_data.csv")
# Remove unique identification column
data <- data[, !(names(data) %in% c("X", "seqn"))]
summary(data)
```

## Categorical Data

- Response: event
```{r}
table(data$event)
```

- gender
Assign all female to 0, and keep male be 1. So, now we have, 0-female and 1-male.
```{r}
data$gender[data$gender == 2] <- 0
mosaicplot(table(data$gender, data$event), color=c("blue", "red"), main="Mosaic Plot for Event vs Gender", xlab = 'Gender', ylab = 'Event')
```

- ethnic1
Except Non-Hispanic black be black and Non-Hispanic white be white, we assign all other categories be the other.
```{r}
data <- data %>%
  mutate(ethnic1 = case_when(
    ethnic1 == 1 ~ "Other",
    ethnic1 == 2 ~ "Other",
    ethnic1 == 3 ~ "White",
    ethnic1 == 4 ~ "Black",
    ethnic1 == 5 ~ "Other",
    TRUE ~ as.character(ethnic1)
  ))
mosaicplot(table(data$ethnic1, data$event), color=c("blue", "red"), main="Mosaic Plot for Event vs Ethnic", xlab = 'Ethnic', ylab = 'Event')
```
- educ
- diabetes
- smoker
NA value replacement. For the NA value for smoker, we use gender, edu, and age be the predictor to form a logistic regression to predict the value.
```{r}
# 3. smoker
data_no_na <- na.omit(data)
table(data$smoker)
# For Categorical response, first Categorize the result to (0-No smoke; 1-smoke)
# turn 2 into 0, which not only turns data to correct format of data for logistic regression but also more intuitive.
data$smoker[data$smoker == 2] <- 0
smoke_pr <- glm(smoker ~ gender + educ + age, data = data, family = binomial)
# Model evaluation
#summary(smoke_pr)
#data_no_na$smoker[data_no_na$smoker == 2] <- 0
#predicted_smoker <- ifelse(predict(smoke_pr, newdata = data_no_na) >= 0.5, 1, 0)
#mean(data_no_na$smoker == predicted_smoker)
# Missing value replace
rows_with_na <- is.na(data$smoker)
predicted_probabilities <- predict(smoke_pr, newdata = data[rows_with_na, ], type = "response")
data$smoker[rows_with_na] <- ifelse(predicted_probabilities >= 0.5, 1, 0)
# Check Result
sum(is.na(data$smoker))
summary(data$smoker)
```
Plot the data structure
Here we have, 0-no smoke in past 5 years and 1-smoked in past 5 years
```{r}
mosaicplot(table(data$smoker, data$event), color=c("blue", "red"), main="Mosaic Plot for Event vs Smoker", xlab = 'Smoker', ylab = 'Event')
```


## Numerical Data

- age
```{r}
hist(data$age)
age_model <- glm(event ~ age, data = data, family = binomial, )
summary(age_model)
```

- sleep.hrs
NA value replacement.
```{r}
# According to the correlation, use the highest possible predictors
sleep_fit <- lm(sleep.hrs ~ gender+age+ethnic1+educ, data)
summary(sleep_fit)
# Based on first try, use the most significant predictors
sleep_fit1 <- lm(sleep.hrs ~ gender+ethnic1, data)
summary(sleep_fit1)
# Evaluate Model Performance
# Predicted values from the model
#predicted_sleep <- predict(sleep_fit1, newdata = data_no_na)
# Mean Absolute Error (MAE)
#mae <- mean(abs(data_no_na$sleep.hrs - predicted_sleep)); mae
# Mean Squared Error (MSE)
#mse <- mean((data_no_na$sleep.hrs - predicted_sleep)^2); mse
# Root Mean Squared Error (RMSE)
#rmse <- sqrt(mse); rmse
# R-squared (R²)
#rsquared <- 1 - (sum((data_no_na$sleep.hrs - predicted_sleep)^2) / sum((data_no_na$sleep.hrs - mean(data_no_na$sleep.hrs))^2)); rsquared
# Replace the missing value using the predict model
# Find rows with missing values
rows_with_na <- is.na(data$sleep.hrs)
# Select the rows with missing values
data_sleep_missing <- data[rows_with_na, ]
# Replace the missing values with the predicted value by linear regression model
data$sleep.hrs[rows_with_na] <- predict(sleep_fit1, newdata = data_sleep_missing)
# Check Result
sum(is.na(data$sleep.hrs))
summary(data$sleep.hrs)
```

```{r}
hist(data$sleep.hrs)
sleep_model <- glm(event ~ sleep.hrs, data = data, family = binomial)
summary(sleep_model)
```


- bmi
```{r}
data$bmi.cat <- cut(data$bmi, breaks=c(-Inf, 18.5, 24.9, 29.9, Inf), labels=c("Underweight","Normal","Overweight", "Obese"))
table(data$bmi.cat)
round(prop.table(table(data$bmi.cat))*100, 4)
data %>%
  filter(bmi.cat == 'Underweight') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(bmi.cat == 'Normal') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(bmi.cat == 'Overweight') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(bmi.cat == 'Obese') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )
```

```{r}
mosaicplot(table(data$bmi.cat, data$event), color=c("blue", "red"), main="Mosaic Plot for Event vs Smoker", xlab = 'Smoker', ylab = 'Event')
```


NA value replacement. Due to the model performances terrible, we choice the median value to replace the NA value.
```{r}
# According to the correlation, use the highest possible predictors
bmi_model = lm(bmi ~ gender + ethnic1, data)
summary(bmi_model)
# Based on first try, use the most significant predictors
bmi_model1 = lm(bmi ~ age + gender + ethnic1, data)
summary(bmi_model1)
# Evaluate Model Performance
# Predicted values from the model
predicted_bmi <- predict(bmi_model1, newdata = data_no_na)
# Mean Absolute Error (MAE)
mae <- mean(abs(data_no_na$bmi - predicted_bmi)); mae
# Mean Squared Error (MSE)
mse <- mean((data_no_na$bmi - predicted_bmi)^2); mse
# Root Mean Squared Error (RMSE)
rmse <- sqrt(mse); rmse
# R-squared (R²)
rsquared <- 1 - (sum((data_no_na$bmi - predicted_bmi)^2) / sum((data_no_na$bmi - mean(data_no_na$bmi))^2)); rsquared
# Have terrible result in model, so we decided to use the median value to replace the missing value
rows_with_na <- is.na(data$bmi)
data$bmi[rows_with_na] <- median(data_no_na$bmi)
# Check Result
sum(is.na(data$bmi))
summary(data$bmi)
```

```{r}
hist(data$bmi)
bmi_model <- glm(event ~ bmi, data = data, family = binomial)
summary(bmi_model)
```
