---
title: "STSCI5160 - Prelim 4"
author: "ghp_CyO0i6oldLed1ZX3cTBu8MBXSRRE8z4hVbvl"
date: "4043-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Library Imports
```{R}
library(tidyverse)
```

# Data Analyze
```{r}
# Load the CSV file
data <- read.csv("final_cardiac_data.csv")
data_clean = read.csv("final_cardiac_data.csv")
# Data Overview
head(data)
str(data)

# Find all NA value
data[!complete.cases(data), ]

# Find column with NA
columns_with_na <- colSums(is.na(data)) > 0; columns_with_na
```

Question 1:
how to deal with all those missing value

```{r}
hist(data$gender)
hist(data$age)
hist(data$ethnic1)
hist(data$educ)
hist(data$sleep.hrs)
hist(data$diabetes)
hist(data$smoker)
hist(data$bmi)
```

### Handling BMI Missing Values

```{R}
data <- select(data, -c("X", "seqn"))

BMI_NA = sum(is.na(data$bmi)) # 46 missing values
length(data$bmi)
BMI_perc_NA = round((sum(is.na(data$bmi))/length(data$bmi))*100,4)
# only 46/1910 (~4%) of the data in the bmi column is missing
hist(data$bmi)
```

Upon preliminary exploration of the data, there are 46 missing values. This accounts for 4% of the data in the BMI column.

There seem to be outliers in the BMI and a heavy right skew for the distribution of this data. Using something like mean imputation to replace the missing values may not be the best way to represent the data. We will use median imputation instead to replace the 4% of values that are missing. 

```{R}
# median imputation
data_clean$bmi <- ifelse(is.na(data$bmi), median(data$bmi), data$bmi)
hist(data_clean$bmi, col = "lightblue", main = "BMI Distribution", xlab = "BMI")
```

```{R}
summary(data)
```
```{R}
# Remove all na value, and find the relation between each variable
data <- data[, !(names(data) %in% c("X", "seqn"))]
data_no_na <- na.omit(data)
cor_matrix <- cor(data_no_na)
heatmap(cor_matrix)
pairs(data_no_na)
```

```{R}
## code for model to fill in missing values (for ryan )
bmi_model = lm(bmi ~ educ + age + gender + ethnic1, data = na.omit(data))
summary(bmi_model) # R^4 = 0.01365
# educ and age are insignificant predictors, so we drop them from the model
bmi_model = lm(bmi ~ gender + ethnic1, data = na.omit(data))
summary(bmi_model) 

# Find rows with missing values
rows_with_na <- is.na(data$bmi)
# Select the rows with missing values
data_missing <- data[rows_with_na, ]
# Use the linear regression model to predict 'bmi'
predicted_values <- predict(bmi_model, newdata = data_missing)
# Replace the missing values with the predicted values in the original data frame
data$bmi[rows_with_na] <- predicted_values
```




```{R}
## EXTRA -- WORK IN PROGRESS
ggplot(data_clean, aes(x = bmi, group = factor(event))) +
  geom_boxplot() +
  ggtitle("BMI Distribution by Cardiac Arrest") +
  xlab("BMI") +
  ylab("Cardiac Event")
```


--- WORK IN PROGRESS ---

## Description of characteristics of subjects and written description




```{R}
## EXTRA -- WORK IN PROGRESS
ggplot(data_clean, aes(x = bmi, group = factor(event))) +
  geom_boxplot() +
  ggtitle("BMI Distribution by Cardiac Arrest") +
  xlab("BMI") +
  ylab("Cardiac Event")
```


--- WORK IN PROGRESS ---

## Description of characteristics of subjects and written description

```{R}
# RESPONSE VARABLE BREAKDOWN
prop.table(table(data$event))
```

According to the above results 44.46% of the respondents did have a doctor tell them they have experienced one of the following cardiac events: congestive heart failure, angina, coronary heart disease, hypertension, MI (“heart attack”) or stroke. On the other hand, 57.54% of the respondents do not fall under this description. These results being split relatively evenly signals to us that we do not need to perform any special reweighting to these classes. The logistic regression model should be able to produce valid and applicable results since the data is relatively balanced between those who did not experience a cardiac episode (0) and those who have experienced a cardiac episode. 

```{R, Table 1}
# Table 1: Descriptive statistics for cardiac event predictors

## Categorical Predictors

table(data$gender)
round(prop.table(table(data$gender))*100, 4)

## TO-DO: age has a lot of bins right now...
table(data$age)
round(prop.table(table(data$age))*100, 4)

## recoded as per the guidelines
data <- data %>%
  mutate(ethnic1 = case_when(
    ethnic1 == 1 ~ "White",
    ethnic1 == 2 ~ "Other",
    ethnic1 == 3 ~ "White",
    ethnic1 == 4 ~ "Black",
    ethnic1 == 5 ~ "Other",
    TRUE ~ as.character(ethnic1)
  ))
table(data$ethnic1)
round(prop.table(table(data$ethnic1))*100, 2)

table(data$educ)
round(prop.table(table(data$educ))*100, 4)

table(data$diabetes)
round(prop.table(table(data$diabetes))*100, 2)

table(data$smoker)
round(prop.table(table(data$smoker))*100, 2)

# recoded bmi
data$bmi.cat <- cut(data$bmi, breaks=c(-Inf, 18.5, 24.9, 29.9, Inf), labels=c("Underweight","Normal","Overweight", "Obese"))
table(data$bmi.cat)
round(prop.table(table(data$bmi.cat))*100, 4)

```


```{R, Table 2}
summary(data$sleep.hrs)
```


```{R, % of Event}
table(data$gender)
round(prop.table(table(data$gender))*100, 4)

data %>%
  filter(gender == 1) %>%  # males
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(gender == 2) %>%  # females
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

### 

data %>%
  filter(ethnic1 == 'White') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )


data %>%
  filter(ethnic1 == 'Black') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )


data %>%
  filter(ethnic1 == 'Other') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

###

data %>%
  filter(bmi.cat == 'Underweight') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(bmi.cat == 'Normal') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(bmi.cat == 'Overweight') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(bmi.cat == 'Obese') %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

###

data %>%
  filter(educ == 1) %>%  
  summarise(
    count_0 = sum(event == 0),
    count_1 = sum(event == 1),
    prop_0 = round(count_0 / n(), 4)*100,
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(educ == 2) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(educ == 3) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(educ == 4) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )

data %>%
  filter(educ == 5) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )

###

data %>%
  filter(diabetes == 0) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )


data %>%
  filter(diabetes == 1) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )

###

data %>%
  filter(smoker == 1) %>%  
  summarise(
    count_0 = sum(event == 0),
    prop_0 = round(count_0 / n(), 4)*100,
    count_1 = sum(event == 1),
    prop_1 = round(count_1 / n(), 4)*100
  )

```
```{R, Two-way contingency analysis of variables (one at a time) and their association with outcome variable}

###

contingency_table <- table(data$event, data$gender)
chisq.test(contingency_table)

###

data$ethnic1.cat <- factor(data$ethnic1.cat)
contingency_table <- table(data$event, data$ethnic1.cat)
chisq.test(contingency_table)

###

data$bmi.cat <- factor(data$bmi.cat)
contingency_table <- table(data$event, data$bmi.cat)
chisq.test(contingency_table)

###
data$educ<- factor(data$educ)
contingency_table <- table(data$event, data$educ)
chisq.test(contingency_table)

###

data$diabetes<- factor(data$diabetes)
contingency_table <- table(data$event, data$diabetes)
chisq.test(contingency_table)

###

data$smoker<- factor(data$smoker)
contingency_table <- table(data$event, data$smoker)
chisq.test(contingency_table)

```









